Confirmation of Wide-Field Signatures in Redshifted 21~cm Power Spectra using Murchison Widefield Array Observations

### 2015-04-25 ms.tex
Update ms.bbl included in ms.tex
From 03f54dc:
============

In ms.tex,

(*) Update ms.bbl included in ms.tex
Make changes suggested by Melanie
From 6501829:
============

In ms.tex,

(*) Make changes suggested by Melanie
(*) Minor changes in Introduction
(*) Chnage GLEAM author to Wayth

% This analysis confirms recent predictions of certain key signatures in the delay spectra of wide-field radio interferometer measurements. While probing H~{\sc i} in the epoch of reionization (EoR) through modeling of delay spectra of measurements between antenna pairs, it has recently emerged that the foreground structure, the nature of wide-field measurements and the antenna aperture imprint a characteristic {\it pitchfork}-shaped signature. This feature is characterized by enhanced power from foreground emission in Fourier modes mapped to regions near the horizon, and is significant for EoR studies because Fourier modes considered to be most sensitive for the cosmological H~{\sc i} signal are very susceptible to contamination from these regions. Thus it forms a critical input to design and analysis choices of future instruments such as the Hydrogen Epoch of Reionization Array and the Square Kilometre Array. By improving Murchison Widefield Array data sensitivity through coherent averaging of 12 independent snapshots aligned in local sidereal time across different observing nights, we provide the first confirmation of the prediction with a signal-noise ratio $>10$.

@@

% The primary challenge to detection of cosmological H~{\sc i} from the EoR comes from continuum emission from Galactic and extragalactic foreground objects, relative to which the desired signal is $\sim 10^4$ times weaker \citep[see, e.g.,][]{dim02,zal04,fur06,ali08,ber09,ber10,gho12}. Since the foregrounds and the signal have inherent differences in spatial isotropy and spectral smoothness \citep{mor04,mor06,bow09,liu11,par12b,dil13,pob13}, a detailed characterization of foreground emission has become essential \citep{bow09,liu09,dat10,liu11,mor12,tro12,pob13,dil14,liu14a,liu14b,thy13,thy15}.

The primary challenge to detection of cosmological H~{\sc i} from the EoR arises from continuum emission from Galactic and extragalactic foreground objects, relative to which the desired signal is $\sim 10^4$ times weaker. But the inherent differences in spatial isotropy and spectral smoothness can be exploited to extract the cosmological signal from foreground contamination \citep[see, e.g.,][]{dim02,dim04,zal04,fur04,mor04,san05,fur06,mcq06,mor06,wan06,gle08}. Thus, a detailed characterization of foreground emission has become essential \citep{ali08,bow09,liu09,ber09,ber10,dat10,liu11,gho12,mor12,par12b,tro12,pob13,dil13,dil14,liu14a,liu14b,thy13,thy15}.

% A recent study by \citet{thy15} using instrument and all-sky foreground models has shed new light on the significant role played by the interplay between foreground emission and the nature of wide-field measurements, typical of most modern EoR experiments. It was noted to cause significant contamination in Fourier modes considered to be most sensitive for the cosmological H~{\sc i} signal. It has been shown that a careful design of antenna aperture can significantly mitigate this contamination. And an optimal weighting of foreground-contaminated Fourier modes may be required to extract the signal with maximum sensitivity. Thus, knowledge of such detailed foreground signatures becomes essential for design and analysis choices of future instruments such as HERA and SKA. This paper presents a follow-up analysis of that study. By using a larger volume of MWA data obtained independently over different nights pointed at the same region of sky, we confirm with high significance the presence of certain key characteristics (referred to as the {\it pitchfork} signature) of wide-field measurements predicted in the preceding study.

Our recent study \citep[][hereafter referred to as Paper~I]{thy15} used all-sky foreground and instrument models for the first time in order to simulate actual EoR experiments more accurately than previous studies. Surprisingly, we found that foreground emission outside the primary beam field of view caused the most significant contamination of the cosmological H~{\sc i} power spectrum Fourier modes that are considered most sensitive for detecting the signal in delay spectrum based analyses. This contamination is the result of the interplay between foreground emission, particularly diffuse Galactic emission, and the wide-field properties typical of EoR instruments. Our simulations predicted that delay spectra from the MWA and other experiments should exhibit a characteristic ``pitchfork'' appearance with local maxima near the horizon delay limits, in addition to at the primary lobe region.

@@

% Detailed modeling presented in Paper~I has important implications for future instrument design and analysis algorithms:
% \begin{enumerate}
% \item positioned adjacent to the Fourier modes most sensitive to the cosmological signal, the {\it pitchfork} feature is a leading source of contamination in these modes, and 
% \item of the cases considered, a dish-shaped antenna was found to yield the most desirable response from a foreground contamination viewpoint due to a relatively efficient suppression of sensitivity away from the primary field of view. 
% \end{enumerate}

@@

\begin{figure}[htb]
\centering
% \includegraphics[width=\linewidth]{multi_baseline_CLEAN_fhd_avg_visibilities_amplitudes_185.0_MHz_30.7_MHz.eps}
\includegraphics[width=\linewidth]{multi_baseline_fhd_sim_visibilities_amplitudes_comparison_185.0_MHz_30.7_MHz.eps}
\caption{Delay power spectra obtained from a single snapshot ({\it top}), by averaging 12 snapshots of LST aligned MWA data ({\it middle}), and from modeling with no thermal noise added ({\it bottom}). The $x$-axis, denoted by $|\boldsymbol{b}|$ (and $k_\perp$), represents angular (and spatial) scales in the plane of the sky while the $y$-axis, shown in $\tau$ and $k_\parallel$, denotes the spatial scales along the line of sight. White dotted lines are the horizon delay limits. Dynamic range in the delay power spectra of MWA data has increased by a factor $\sim 10$ after averaging (middle) relative to that in a single snapshot (top). Power near the horizon limits caused by wide-field effects are prominent. Faint horizontal features at $\tau=\pm 0.78\,\mu$s are visible due to effective lowering of thermal fluctuations and are the response to periodic coarse band edge flagging of MWA data every 1.28~MHz. \label{fig:delay-spectra}}
\end{figure}

@@

We focus on the peaks near the horizon limits in the data. Typically, the power near the negative horizon limit is seen with a signal-noise ratio (SNR) $\sim$~10-100, while that around the positive horizon limit is $\sim$~100-1000. % The models of Paper~I have noted that the foreground power near the horizon limits is due to the nature of wide-field measurements, and is predominantly composed of diffuse emission especially on baseline lengths $\lesssim 100$~m. 
Based on the morphological agreement between the data and the models, we conclude the features noted in our current analysis are a robust detection of the {\it pitchfork} signature predicted in our earlier study.

@@

Precise modeling is thus required to gain a complete understanding of the characteristics of the cosmological signal and the foregrounds. Extending the footprint of surveys matching the frequency and angular resolution of observation such as the MWA Commissioning Survey \citep[MWACS;][]{hur14} and the Galactic and Extragalactic MWA Survey (GLEAM; Hurley-Walker et al. in prep.) to cover the entire hemisphere, and a detailed model of the instrument will further improve the accuracy of our models. In our earlier study, we proposed a selective flagging of data on different antenna spacings that can potentially mitigate foreground contamination by two orders of magnitude. Following the confirmation presented here, efforts are underway to incorporate this proposed foreground mitigation technique into the MWA data analysis algorithms.

@@

% Fluctuations of redshifted H~{\sc i} from the reionization epoch are extremely faint relative to the radio foregrounds. This poses perhaps the greatest challenge to EoR experiments and thus merits a detailed characterization in order to separate them from the desired cosmological H~{\sc i} signal.

% In a recent study, through the use of all-sky foreground and instrument models, the role of wide-field nature of radio interferometric measurements, hitherto unpredicted, has emerged. Referred to as the {\it pitchfork} signature in the delay spectral domain, it is characterized by significant foreground power in the primary field of view and an enhancement in power near the horizon. The latter is due to the highly nonlinear mapping between geometric delays and subtended solid angles, besides an increase in sensitivity to larger scales caused by foreshortening of baseline lengths along these directions. Owing to its localization in the delay domain, it is a leading cause of contamination in Fourier modes most sensitive to the EoR H~{\sc i} signal.

% Knowledge of these characteristics weigh heavily on choices in instrument design like antenna aperture and in data analysis such as optimal weighting schemes for future instruments such as HERA and SKA. We report the first robust detection of the {\it pitchfork} signature (which could not be confirmed earlier in single MWA snapshots due to large thermal fluctuations) with a signal-noise ratio $>10$ made possible by coherent averaging of 12 independent LST-aligned snapshots from MWA observations.

@@

>>>>

@@

The primary challenge to detection of cosmological H~{\sc i} from the EoR arises from continuum emission from Galactic and extragalactic foreground objects, relative to which the desired signal is $\sim 10^4$ times weaker. However, the inherent differences in spatial isotropy and spectral smoothness can be exploited to extract the cosmological signal from foreground contamination \citep[see, e.g.,][]{dim02,dim04,zal04,fur04,mor04,san05,fur06,mcq06,mor06,wan06,gle08}. Thus, a detailed characterization of foreground emission has become essential \citep{ali08,bow09,liu09,ber09,ber10,dat10,liu11,gho12,mor12,par12b,tro12,pob13,dil13,dil14,liu14a,liu14b,thy13,thy15}.

Our recent study \citep[][hereafter referred to as Paper~I]{thy15} used all-sky foreground and instrument models for the first time in order to simulate actual EoR experiments more accurately than previous studies. Surprisingly, we found that foreground emission outside the primary beam field of view caused the most significant contamination of Fourier modes considered most sensitive for detecting the cosmological H~{\sc i} signal in delay spectrum based analyses. This contamination is the result of the interplay between foreground emission, particularly diffuse Galactic emission, and the wide-field properties typical of EoR instruments. Our simulations predicted that delay spectra from the MWA and other experiments should exhibit a characteristic ``pitchfork'' appearance with local maxima near the horizon delay limits, in addition to at the primary lobe region.

@@

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{multi_baseline_fhd_sim_visibilities_amplitudes_comparison_185.0_MHz_30.7_MHz.eps}
\caption{Delay power spectra obtained from a single snapshot ({\it top}), by averaging 12 snapshots of LST aligned MWA data ({\it middle}), and from modeling with no thermal noise added ({\it bottom}). The $x$-axis, denoted by $|\boldsymbol{b}|$ (and $k_\perp$), represents angular (and spatial) scales in the plane of the sky while the $y$-axis, shown in $\tau$ and $k_\parallel$, denotes the spatial scales along the line of sight. White dotted lines are the horizon delay limits. Dynamic range in the delay power spectra of MWA data has increased by a factor $\sim 10$ after averaging (middle) relative to that in a single snapshot (top). Power near the horizon limits caused by wide-field effects are prominent. Faint horizontal features at $\tau=\pm 0.78\,\mu$s are visible due to effective lowering of thermal fluctuations and are the response to periodic coarse band edge flagging of MWA data every 1.28~MHz. \label{fig:delay-spectra}}
\end{figure}

@@

We focus on the peaks near the horizon limits in the data. Typically, the power near the negative horizon limit is seen with a signal-noise ratio (SNR) $\sim$~10-100, while that around the positive horizon limit is $\sim$~100-1000. Based on the morphological agreement between the data and the models, we conclude the features noted in our current analysis are a robust detection of the {\it pitchfork} signature predicted in our earlier study.

@@

Precise modeling is thus required to gain a complete understanding of the characteristics of the cosmological signal and the foregrounds. Extending the footprint of surveys matching the frequency and angular resolution of observation such as the MWA Commissioning Survey \citep[MWACS;][]{hur14} and the Galactic and Extragalactic MWA Survey (GLEAM; Wayth et al. submitted) to cover the entire hemisphere, and a detailed model of the instrument will further improve the accuracy of our models. In our earlier study, we proposed a selective flagging of data on different antenna spacings that can potentially mitigate foreground contamination by two orders of magnitude. Following the confirmation presented here, efforts are underway to incorporate this proposed foreground mitigation technique into the MWA data analysis algorithms.

@@

@@

% \bibitem[{{Hurley-Walker} {et~al.}(2014){Hurley-Walker}, {Morgan}, {Wayth},
%   {Hancock}, {Bell}, {Bernardi}, {Bhat}, {Briggs}, {Deshpande}, {Ewall-Wice},
%   {Feng}, {Hazelton}, {Hindson}, {Jacobs}, {Kaplan}, {Kudryavtseva}, {Lenc},
%   {McKinley}, {Mitchell}, {Pindor}, {Procopio}, {Oberoi}, {Offringa}, {Ord},
%   {Riding}, {Bowman}, {Cappallo}, {Corey}, {Emrich}, {Gaensler}, {Goeke},
%   {Greenhill}, {Hewitt}, {Johnston-Hollitt}, {Kasper}, {Kratzenberg},
%   {Lonsdale}, {Lynch}, {McWhirter}, {Morales}, {Morgan}, {Prabu}, {Rogers},
%   {Roshi}, {Shankar}, {Srivani}, {Subrahmanyan}, {Tingay}, {Waterson},
%   {Webster}, {Whitney}, {Williams}, \& {Williams}}]{hur14}
% {Hurley-Walker}, N., {Morgan}, J., {Wayth}, R.~B., {et~al.} 2014, \pasa, 31, 45
